---
title: "breaks in histograms (and count vs density)"
author: "Alan Cobo-Lewis"
date: "2025-02-12"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load library}
library(tidyverse)
```

## with default `geom_histogram()`, first 7 bins span just over 0 to just over 5000, so that's a binwidth of about 714
```{r}
ggplot(diamonds, aes(x=price)) + geom_histogram()
```

## Try binwidths of exactly 700 from 0 to 7000, then one big bin from 7000 to 20000
```{r}
ggplot(diamonds, aes(x=price)) +
  geom_histogram(
    breaks=c(0,700,1400,2100,2800,3500,4200, 4900, 5600,6300,7000, 20000)
  ) +
  labs(title="Uneven break spacing for sure, but what's wrong here?")
```

## Make the same weird plot by *explicitly* specifying the y aesthetic for `geom_histogram()` and `stat_bin()`. Notice that we have to use the `after_stat()` wrapper because the `count` variable doesn't exist in the `diamonds` dataframe--it's *calculated* by the `stat_bin()` function
```{r}
ggplot(diamonds, aes(x=price)) +
  geom_histogram(
    aes(y=after_stat(count)),
    breaks=c(0,700,1400,2100,2800,3500,4200, 4900, 5600,6300,7000, 20000)
  ) +
  labs(title="Still uneven break spacing, but what's still wrong?")
```

## Now change from default `aes(y=after_stat(count))` to more appropriate `aes(y=after_stat(density))`
```{r}
ggplot(diamonds, aes(x=price)) +
  geom_histogram(
    aes(y=after_stat(density)),
    breaks=c(0,700,1400,2100,2800,3500,4200, 4900, 5600,6300,7000, 20000)
  ) +
  labs(title='Makes more sense with density (count per unit x variable) on y axis')
```

## And that's why the alternative is called `geom_density()`
```{r}
ggplot(diamonds, aes(x=price)) +
  geom_histogram(
    aes(y=after_stat(density)),
    breaks=c(0,700,1400,2100,2800,3500,4200, 4900, 5600,6300,7000, 20000)
  ) +
  geom_density(color='red') +
  labs(title='Same y axis works for geom_density() and for\ngeom_histogram(aes(y=after_stat(density)))')
```
