---
title: "anscombe"
author: "Alan Cobo-Lewis"
date: "2025-01-29"
output: html_document
---

```{r setup, include=FALSE}
library("knitr")     # for rendering the RMarkdown file
library("tidyverse") # for plotting (and many more cool things we'll discover later)

knitr::opts_chunk$set(echo = TRUE)
```

The data are in the `datasets` package, which is already loaded into R by default. (If it weren't, you could always do `library(datasets)` -- after doing `install.packages(datasets)` if you hadn't done it before). The data are in the dataframe `anscombe` (which doesn't follow the `df.` naming convention that our text does--sorry).

```{r look at the data}
anscombe
```

The `anscombe` dataframe is "wide". It's easier to work with in the `tidyverse` if its pivoted into "long" format. We'll cover these kind of data manipulation in a subsequent module, but at least let's adopt our friendly `df.` naming convention when we make this new long object:

```{r stackit!, echo=TRUE}
df.long.anscombe <- anscombe %>%
  pivot_longer(
    everything(),
    cols_vary = "slowest",
    names_to = c(".value", "set"),
    names_pattern = "(.)(.)"
  )
df.long.anscombe
```

We can replicate the plots:

```{r plotit!, echo=TRUE}
ggplot(df.long.anscombe, aes(x=x, y=y)) +
  geom_point() +
  facet_wrap(~set)
```

Alan thought the Pearson correlation coefficient (*r*) was about 0.7, but it's actually around 0.8. Notice that the means, SDs, and *r*s are the same across all four sets:

```{r statit!, echo=TRUE}
df.long.anscombe %>%
  group_by(set) %>%
  summarise(
    count=n(),
    meanx=mean(x),
    meany=mean(y),
    sdx=sd(x),
    sdy=sd(y),
    r=cor(x,y)
  )
```
